---
title: "Master Seminar. Case 3 Monte Carlo simulatio"
subtitle: "Monte Carlo Simulation of Heterogeneous Treatment Effects"
author: "Roman Rakov"
date: "05/24"
output: 
  html_notebook:
    toc: true
    toc_float: true
    code_folding: show
---

```{r message=FALSE, warning=FALSE}
library(plm)
library(dplyr)
library(grf)
library(causalDML)
library(tidyverse)
library(patchwork)
library(estimatr)
library(np)
library(crs)
library(gridExtra)
library(scales) 
```

<br>

## Non-parametric heterogeneity

**DGP**: zero ATE and highly nonlinear effect heterogeneity

-   $p=10$ independent covariates $X_1,...,X_k,...,X_{10}$ drawn from a uniform distribution: $X_k \sim uniform(-\pi,\pi)$

-   Propensity score: $W \sim Bernoulli(\underbrace{\Phi(sin(X_1))}_{e(X)})$, where $\Phi(\cdot)$ is the standard normal cdf

-   Potential outcome of controls: $Y(0) = \underbrace{cos(X_1+1/2\pi)}_{m_0(X)} + u_i \times 0.5 X^2 + \varepsilon$, with $\varepsilon \sim N(0,5)$

-   Potential outcome of treated is $Y(1) = \underbrace{sin(X_1)}_{m_1(X)} + u_i \times 0.5 X^2 + \varepsilon$, with $\varepsilon \sim N(0,5)$

-   The treatment effect function is $\tau(X) = sin(X_1) - cos(X_1+1/2\pi)$

-   Each unit has its unit-specific confounding, which complicates matters: $u_i \sim N(0,5)$

This simulation is a nice exercise for the Kernel regression.

```{r}
# Set parameters
n <- 1000
p <- 10

# Illustration
x <- matrix(runif(n*p, -pi, pi), ncol = p)
e <- function(x) { pnorm(sin(x)) }
m1 <- function(x) { sin(x) }
m0 <- function(x) { cos(x + 1/2 * pi) }
tau <- function(x) { m1(x) - m0(x) }
w <- rbinom(n, 1, e(x[,1]))
y <- w * m1(x[,1]) + (1 - w) * m0(x[,1]) + rnorm(n, 0, 1)

# Function to create a custom theme with transparent background
custom_theme <- function(base_size = 12) {
  theme_minimal(base_size = base_size) +
    theme(
      text = element_text(size = base_size + 2),
      axis.title = element_text(face = "bold"),
      plot.title = element_text(face = "bold", size = base_size + 4),
      panel.background = element_rect(fill = alpha("lightgray", 0.5)),  
      panel.grid.major = element_line(color = "white"),
      panel.grid.minor = element_line(color = "white"),
      legend.position = "none")}

# pdf("case_3_plot.pdf", width = 10, height = 6)  # Dimensions in inches

# Create the plots
g1 <- data.frame(x = c(-pi, pi)) %>%
  ggplot(aes(x)) + 
  stat_function(fun = e, linewidth = 1, aes(colour = "e")) + 
  labs(x = "Propensity score along X1", y = "e") +
  scale_colour_manual(values = c("e" = "darkred")) +
  custom_theme() +
  theme(legend.position = "none")

g2 <- data.frame(x = c(-pi, pi)) %>%
  ggplot(aes(x)) + 
  stat_function(fun = m1, linewidth = 1, aes(colour = "Y1")) + 
  stat_function(fun = m0, linewidth = 1, aes(colour = "Y0")) + 
  labs(x = "Potential outcomes along X1", y = "Y") +
  scale_colour_manual(values = c("Y1" = "blue", "Y0" = "red")) +
  custom_theme() +
  geom_text(x = 0, y = m1(0), label = "Y(1)", vjust = -2, colour = "blue", size = 5, hjust = 0) +
  geom_text(x = 0, y = m0(0), label = "Y(0)", vjust = 2.5, colour = "red", size = 5, hjust = 0)

g3 <- data.frame(x = c(-pi, pi)) %>%
  ggplot(aes(x)) + 
  stat_function(fun = tau, linewidth = 1, aes(colour = "tau")) + 
  labs(x = "Treatment effect along X1", y = expression(tau)) +
  scale_colour_manual(values = c("tau" = "darkgreen")) +
  custom_theme()

# Arrange plots vertically
grid.arrange(g1, g2, g3, nrow = 3)
# dev.off()
```

Simulation:

```{r warning=FALSE}
set.seed(123)

num_units    <- 5
obs_per_unit <- 1000
unit_effects <- rnorm(num_units, mean = 0, sd = 10) 
unit_covar   <- runif(num_units,-pi,pi)


# Create a data frame to hold the panel data
panel_data <- expand.grid(unit = 1:num_units, obs = 1:obs_per_unit)

# Generate covariates X1, X2, X3
panel_data <- panel_data %>%
  mutate(X1 = runif(n(),-pi,pi),
         X2 = runif(n(),-pi,pi),
         X3 = runif(n(),-pi,pi)) 

# Generate treatment dummy (randomly assigned)
panel_data <- panel_data %>%
  mutate(treatment = rbinom(n(),1,pnorm(sin(X1))))

# Generate the outcome variable with unit effects, treatment effect, and random noise
panel_data <- panel_data %>%
  group_by(unit) %>%
  mutate(outcome = unit_effects[unit]*0.5*X2^2 + treatment*sin(X1) + (1-treatment)*cos(X1+1/2*pi) + rnorm(n(),0,5)) %>%
  ungroup()

# Convert data to panel data structure
# panel_data <- pdata.frame(panel_data, index = c("unit", "obs"))

# Calculate the true average treatment effect
true_ate <- mean(panel_data$treatment*sin(panel_data$X1) + (1-panel_data$treatment)*cos(panel_data$X1+1/2*pi)) 
print(paste("True Average Treatment Effect:", true_ate))

pooled_ols <- lm(outcome ~ treatment + X1 + X2 + X3 + X1*treatment, data = panel_data)
fixed_effects <- plm(outcome ~ treatment + X1 + X2 + X3 + X1*treatment, data = panel_data, model = "within")

print(summary(pooled_ols))    # Pooled OLS Results
print(summary(fixed_effects)) # Fixed Effects Estimation Results
```

<br>

### CATE estimation

Raw data:

```{r}
W <- panel_data$treatment 
Y <- panel_data$outcome
X <- panel_data %>%
  select(X1, X2, X3)

# Run the main function that outputs nuisance parameters, APO and ATE
forest <- create_method("forest_grf", args = list(tune.parameters = "all", seed = 123))
aipw <- DML_aipw(Y, W, X, ml_w = list(forest), ml_y = list(forest), cv = 3)

# temp_ols <- lm_robust(aipw$ATE$delta ~ ., data = X)
# print(summary(temp_ols))
```

Now, instead of running an OLS (*this time heterogeneity is non-linear!*), we use this pseudo-outcome $\tilde{Y}_{ATE}$ as a dependent variable with $X_1$ as the only covariate. We cross-validate the bandwidth of the kernel regression, run the estimation with the `np` package and plot the estimated curve:

```{r message=FALSE}
kernel_raw_x1 = kr_cate(aipw$ATE$delta, X[, 1])

#pdf("kernel_raw_x1.pdf", width = 10, height = 6.5)  # Dimensions in inches
plot(kernel_raw_x1,z_label = 'X1 value')
#dev.off()
```

```{r}
spline_raw = spline_cate(aipw$ATE$delta,X[, 1])

# pdf("spline_raw.pdf", width = 10, height = 6.5)  # Dimensions in inches
print(plot(spline_raw,z_label="X1 value")) 
# dev.off()
```

A lot of noise. Now, let's check other variables:

```{r}
kernel_reg_x2 = kr_cate(aipw$ATE$delta,X[, 2])
plot(kernel_reg_x2)
```

Nothing (?), as it should be. Now try same thing with demeaning:

```{r}
X_demean <- panel_data %>%
  select(X1, X2, X3, unit) %>%
  group_by(unit) %>%
  mutate_at(vars(-group_cols()), ~. - mean(.)) %>%
  ungroup() %>%
  select(-unit)

Y_demean <- panel_data %>%
  select(outcome, unit) %>%
  group_by(unit) %>%
  mutate(outcome = outcome - mean(outcome)) %>%
  ungroup() %>%
  select(-unit)

Y_demean <- as.numeric(unlist(Y_demean))

aipw_demean <- DML_aipw(Y_demean, as.numeric(W), X_demean, ml_w = list(forest), ml_y = list(forest), cv = 3)

# temp_ols_demean <- lm_robust(aipw_demean$ATE$delta ~ ., data = X_demean)
# print(summary(temp_ols_demean))
```

```{r message=FALSE}
kernel_reg_x1 = kr_cate(aipw_demean$ATE$delta,X_demean[, 1])

# pdf("kernel_demean.pdf", width = 10, height = 6.5)  # Dimensions in inches
plot(kernel_reg_x1,z_label = "X1 value")
# dev.off()
```

```{r}
spline_demean = spline_cate(aipw_demean$ATE$delta,X_demean[, 1])

#pdf("spline_demean.pdf", width = 10, height = 6.5)  # Dimensions in inches
print(plot(spline_demean,z_label="X1 value")) 
#dev.off()
```
